\textsc{RILS}-\textsc{ROLS} : Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-11-26 paper_resources/doc/paper.tex
Finished up to 4.1.

	\section{The proposed method}\label{sec:rils}

	Before we give details on our \textsc{RILS}-\textsc{ROLS}  method for solving SR, we will explain two building blocks of \textsc{RILS-ROLS}: 1) iterated local search (ILS) and 2) ordinary least squares (OLS).

@@

Candidate perturbations are being improved by performing OLS coefficient fitting (procedure \texttt{FitOLS}). This means that coefficients in any of linear combinations of current solution are being set by applying  ordinary least square method, already described in Section~\ref{sec:ols}.    \fxnote{Marko: Aco, dodaj ono oko  nelinearnog dijela i pordešavanja koeficijenata koji se traže u predefinisanom skupu vrijednosti...}
After this step, perturbations are usually better suited to given sample data $D_{tr}'$. 
Further, these perturbations are sorted w.r.t. $R^2$ metric in the descending order (line~\ref{line:orderR2}).

@@

\begin{enumerate}
	\item Any node to any of its subtrees (excluding itself). For example, if $(x+y^3)$ is changed to $x$, the perturbation is $(x+y^3)\cdot x/\sqrt{y} \rightarrow x\cdot x/\sqrt{y}$. 
	\item Variable to another variable or constant to variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (y+y^3)\cdot x/\sqrt{y}$.
	\item Variable to unary operation applied to that variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot \ln{x}/\sqrt{y}$.
	\item Unary operation to another unary operation. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot x/\sin{y}$.
	\item Binary operation to another binary operation. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot (x + \sqrt{y})$. 
	\item Variable or constant enter the binary operation with arbitrary variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+x/y^3)\cdot x/\sqrt{y}$. 
\end{enumerate}

@@

\begin{algorithm}
	\hspace*{\algorithmicindent} \textbf{Input}: perturbation $p$, sample training dataset $D_{tr}'$ \\
	\hspace*{\algorithmicindent} \textbf{Output}: local optimum $bp$ in the vicinity of perturbation $p$
	\begin{algorithmic}[1] 
		\Procedure{LocalSearch}{$p$, $D_{tr}'$}
		\State $bp, bp_{fit} \gets p,\ $\texttt{Fitness}$(p,D_{tr}')$ 
		\State $improved \gets true$
		\While{$improved$}
		\State $improved \gets false$
		\State $bp_{candidates} \gets \emptyset$
		\State $bp_{subtrees} \gets $\texttt{SubTrees}($bp$)
		\For{$n \in bp_{subtrees}$}
		\State $n_{candidates} \gets $ \texttt{All1PerturbationsAroundNode}($bp$, $n$)
		\For{$new \in n_{candidates}$}
		\State $new \gets$ \texttt{FitOLS}($new$, $D_{tr}'$)
		\State $new_{fit} \gets$ \texttt{Fitness}($new$, $D_{tr}'$)
		\If{$new_{fit} < bp_{fit}$}
		\State $bp, bp_{fit} \gets new, new_{fit}$
		\State $improved \gets true$
		\EndIf
		\EndFor
		\EndFor
		\EndWhile
		\State \Return $bp$
		\EndProcedure
	\end{algorithmic}
	\caption{Local search procedure.}
	\label{alg:ls}
\end{algorithm}

For a given perturbation $p$, local search systematically explores all 1-perturbations around $p$. It relies on the \emph{best-improvement} strategy, meaning that all 1-perturbations (for all subtrees) are considered. Before checking if the candidate solution ($new$) is better that the actual best $bp$, the OLS coefficient fitting (\texttt{FitOLS}) is performed.

@@

Our \textsc{RILS}--\textsc{ROLS} algorithm is implemented in Python version 3.x. The experiments concerning our method are performed on a machine xx with xx GHz and a memory limit of xx GB in single-threaded mode.  
The following 13 algorithms are compared to our approach: \textsc{AI-Feynman}, \textsc{GOMEA}, \textsc{Afp-Fe}, \textsc{Itea}, \textsc{Afp}, \textsc{Dsr}, \textsc{Operon}, the \textsc{Gplearn} from python package \textsc{gplearn}~\cite{stephens2016genetic}, \textsc{Sbp-Gp}, \textsc{Eplex}, \textsc{Bsr}, \textsc{Feat}, \textsc{Ffx}, \textsc{Mrgp}.

For the (13) competitors, the results are exported from \url{https://cavalab.org/srbench/results/}, reported in~\cite{la2021contemporary}. 
The maximum computation time allowed for each run of \textsc{RILS}-\textsc{ROLS} is set to 1h, i.e., 3600 seconds; note that the same runtime limit was used in case of all (13) competitors, see~\cite{la2021contemporary}.

All    algorithms were run ten times per each problem instance.

>>>>

	\section{The proposed RILS-ROLS method}\label{sec:rils}

 	Our method relies on the following operations: $+$, $-$, $\cdot$, $/$, $\sqrt{x}$, $x^2 $, $\sin$, $\cos$, $\log$, $\exp$, $a^x$, where $a^x$ is only consequently used -- it is never explicitly introduced in the search space. Beside this, the following set of constants enters the search space explicitly: $-1$, $0$, $1$, $2$, $\pi$ and $10$. 
	Before we give details on our   method for solving SR, we will explain two building blocks of \textsc{RILS-ROLS}: 1) iterated local search (ILS) and 2) ordinary least squares (OLS).

@@

Candidate perturbations are being improved by performing OLS coefficient fitting (procedure \texttt{FitOLS}). This means that coefficients in any of linear combinations of current solution are being set by applying  ordinary least square method, already described in Section~\ref{sec:ols}. 
After this step, perturbations are usually better suited to given sample data $D_{tr}'$. 
Further, these perturbations are sorted w.r.t. $R^2$ metric in the descending order (line~\ref{line:orderR2}).

@@

\begin{enumerate}
	\item Any node to any of its subtrees (excluding itself). For example, if $(x+y^3)$ is changed to $x$, the perturbation is $(x+y^3)\cdot x/\sqrt{y} \rightarrow x\cdot x/\sqrt{y}$. 
	\item Constant to variable. For example,  $(1+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot x/\sqrt{y}$.
	\item Variable to unary operation applied to that variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot \ln{x}/\sqrt{y}$.
	\item Unary operation to another unary operation. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot x/\sin{y}$.
	\item Binary operation to another binary operation. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot (x + \sqrt{y})$. 
	\item Variable or constant enter the binary operation with arbitrary variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+x/y^3)\cdot x/\sqrt{y}$. 
\end{enumerate}

@@

\begin{algorithm}
	\hspace*{\algorithmicindent} \textbf{Input}: perturbation $p$, sample training dataset $D_{tr}'$ \\
	\hspace*{\algorithmicindent} \textbf{Output}: local optimum $bp$ in the vicinity of perturbation $p$
	\begin{algorithmic}[1] 
		\Procedure{LocalSearch}{$p$, $D_{tr}'$}
		\State $bp, bp_{fit} \gets p,\ $\texttt{Fitness}$(p,D_{tr}')$ 
		\State $improved \gets true$
		\While{$improved$}
		\State $improved \gets false$
		\State $bp_{candidates} \gets \emptyset$
		\State $bp_{subtrees} \gets $\texttt{SubTrees}($bp$)
		\For{$n \in bp_{subtrees}$}
		\State $n_{candidates} \gets $ \texttt{All1PerturbationsAroundNodeExtended}($bp$, $n$)
		\For{$new \in n_{candidates}$}
		\State $new \gets$ \texttt{FitOLS}($new$, $D_{tr}'$)
		\State $new_{fit} \gets$ \texttt{Fitness}($new$, $D_{tr}'$)
		\If{$new_{fit} < bp_{fit}$}
		\State $bp, bp_{fit} \gets new, new_{fit}$
		\State $improved \gets true$
		\EndIf
		\EndFor
		\EndFor
		\EndWhile
		\State \Return $bp$
		\EndProcedure
	\end{algorithmic}
	\caption{Local search procedure.}
	\label{alg:ls}
\end{algorithm}

For a given perturbation $p$, local search systematically explores extended set of 1-perturbations around $p$. It relies on the \emph{best-improvement} strategy, meaning that all 1-perturbations (for all subtrees) are considered. Before checking if the candidate solution ($new$) is better than the actual best $bp$, the OLS coefficient fitting (\texttt{FitOLS}) takes place.  
The set of used 1-perturbations is increased in comparison to those used previously. Namely, in addition to those six possible types of node changes, the following four are added:

\begin{enumerate}
	\setcounter{enumi}{6}
	\item Any node to any constant or variable. For example, $(x+y^3)\cdot x/\sqrt{y} \rightarrow y \cdot x / \sqrt{y}$ or  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+\pi) \cdot x / \sqrt{y}$.
	\item Any node to unary operation applied to it. For example, $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot x/\ln{\sqrt{y}}$.
	\item Any node to binary operation applied to that node and some variable or constant. For example, $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot x/\sqrt{y} - x$.
	\item Constant to its multiple, where possible multipliers are $\{$0.01, 0.1, 0.2, 0.5, 0.8, 0.9, 1.1, 1.2, 2, 5, 10, 20, 50, 100$\}$. For example, $(1+y^3)\cdot x/\sqrt{y} \rightarrow (1.2+y^3)\cdot x/\sqrt{y}$.
\end{enumerate}

Change under number 10 is very important -- it performs the general coefficient tuning, unlike OLS that considers only coefficients in linear combinations.

@@

Our \textsc{RILS}--\textsc{ROLS} algorithm is implemented in Python 3.9.0. All experiments concerning our method are conducted in the single-core mode, on a PC with Intel i9-9900KF CPU @3.6GHz, 64GB RAM, under Windows 10 Pro OS. The RAM consumption was very small (up to few hundred megabytes) -- beside memory space needed to load input dataset, the only considerable amount is reserved for expression cache.

The following 13 algorithms are compared to our approach: \textsc{AI-Feynman}, \textsc{GOMEA}, \textsc{Afp-Fe}, \textsc{Itea}, \textsc{Afp}, \textsc{Dsr}, \textsc{Operon}, the \textsc{Gplearn} from python package \textsc{gplearn}~\cite{stephens2016genetic}, \textsc{Sbp-Gp}, \textsc{Eplex}, \textsc{Bsr}, \textsc{Feat}, \textsc{Ffx}, \textsc{Mrgp}.

For the (13) competitors, the results are exported from SRBench \url{https://cavalab.org/srbench/results/}, reported in~\cite{la2021contemporary}. 
The maximum computation time allowed for each run of \textsc{RILS}-\textsc{ROLS} is set to 1 hour, while the maximal number of fitness function evaluations is set to 1 million. All other algorithms from SRBench used the time limit that was equal or bellow 8 hours runtime and the same fitness evaluation limit of 1 million (see~\cite{la2021contemporary}). We did not use the full time limit of 8 hours because it was not necessary -- we empirically concluded that there were no improvements after 1 hour \textsc{RILS}-\textsc{ROLS} runtime.

All non-deterministic algorithms (including \textsc{RILS}-\textsc{ROLS}) were run ten times per each problem instance, where each run used different setting of random number generator seed.
